Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-11-28 threshold/draft.Rnw
More examples

@@

Correlation matrix computation can benefit substantially from use of optimized
BLAS Level 3 operations (matrix multiplication), achieving high utilization of
available CPU floating point capacity. The {\tt tcor()} function also benefits
from optimized BLAS routines, but not as much as it mostly makes use of Level 2
operations and Level 3 operations on sequences of smaller problems.

@@

<<echo=FALSE, results=tex>>=
require(xtable)
ans = data.frame(Method=c("tcor","cor"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "BicatYeast results, threshold=0.95", label = "BicatYeast", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

@@

Performance benefits from the {\tt tcor()} function become apparent whith large
data.  The next example uses RPKM-normalized gene expression data measured
across 20,532 genes from 878 breast cancer tumor samples obtained from the
Broad GDAC dashboard\cite{gdac} of the Cancer Genome Atlas data. See the
supplemental code at XXX  (also embedded in this Sweave document) for examples
of efficiently converting the downloaded sample expression data into an R
matrix.

Eight columns with constant zero values were removed from the data in a
pre-processing step, leaving an $878\times 20524$ matrix. The example computes
pairs of gene expression column vectors with correlation values of 0.99
or more. We use the same {\tt tcor} projection dimension of 10 and performance
measurement methodology as used in the first example.

@@

require(xtable)
ans = data.frame(Method=c("tcor","cor"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA BRCA gene expression results (878 samples x 20,542 genes, threshold=0.99)", label = "TCGA1", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

We simulated a data set with a larger number of samples by simply replicating
the rows of the gene expression matrix 27 times to obtain a matrix with 23,706
rows and 20,524 columns, noting that the simulated data have the same rank (of
at most 878) and the same expected number of vector pairs meeting the
correlation threshold as the original data. Each method again computed the same
set of  3,572 pairs of vectors metting the correlation threshold of 0.99, but
the {\tt tcor()} function outperforms the brute force method in both compute
time and memory consumption.

@@

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  # simulated data, replicate rows 27x
}
require(xtable)
ans = data.frame(Method=c("tcor","cor"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA simulated gene expression results (23,706 rows x 20,542 columns, threshold=0.99)", label = "TCGA2", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

@@

XXX Two basic extensions, include functions/options for this in the package.

The main ideas in Algorithm \ref{prune}, although applied to correlation,
easily extend to any distance metric.

Algorithm \ref{prune} can be adapted to find pairs of highly anti-correlated
vectors by changing the distance metric...XXX

@@

%% BROAD GDAC dashboard
\bibitem{gdac}{\url{http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/BRCA/20150821/gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz}}

@@

%% IRLBA implementations
\bibitem{irlbar}
Lewis, Bryan,  \url{https://cran.r-project.org/web/packages/irlba}.

\bibitem{tcor}
Lewis, Bryan,  \url{https://github.com/bwlewis/covar/blob/master/threshold/tcor.R}.

>>>>

Most of the floating point operations occur in step 4 of Algorithm \ref{prune}.
Fortunately the $\ell$ evaluations in step 4 are independent of each other and
can easily be computed in parallel. See the reference R
implementation\cite{sup} for an example.

@@

Correlation matrix computation can benefit substantially from use of optimized
BLAS Level 3 operations (matrix multiplication), achieving high utilization of
available CPU floating point capacity. The {\tt tcor()} function also benefits
from optimized BLAS routines, but not as much, as it mostly makes use of Level
2 operations and Level 3 operations on sequences of smaller problems.

@@

<<echo=FALSE, results=tex>>=
require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "BicatYeast results, threshold=0.95", label = "BicatYeast", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

@@

Performance benefits from the {\tt tcor()} function become apparent with more
data.  The next example uses RPKM-normalized gene expression data measured
across 20,532 genes from 878 breast cancer tumor samples obtained from the
Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}. See the supplemental
code\cite{sup} (also embedded in this Sweave document) for examples of
efficiently converting the downloaded sample expression data into an R matrix.

Eight columns with constant zero values were removed from the data,
leaving an $878\times 20524$ matrix. The example computes
pairs of gene expression column vectors with correlation values of 0.99
or more. We use the same {\tt tcor} projection dimension of 10 and performance
measurement methodology as used in the first example.

@@

require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA BRCA gene expression results (878 samples x 20,542 genes, threshold=0.99)", label = "TCGA1", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

We simulated a data set with a larger number of samples by simply replicating
the rows of the gene expression matrix 27 times to obtain a matrix with 23,706
rows and 20,524 columns, noting that the simulated data have the same rank (of
at most 878) and the same expected number of vector pairs meeting the
correlation threshold as the original data. Each method again computed the same
set of  3,572 pairs of vectors meeting the correlation threshold of 0.99, but
the {\tt tcor()} function outperforms the brute force method in both compute
time and memory consumption.

@@

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  # simulated data, replicate rows 27x
}
require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA simulated gene expression results (23,706 rows x 20,542 columns, threshold=0.99)", label = "TCGA2", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

\subsection{TCGA DNA methylation example}

The following example uses calculated beta values from the Illumina Human
Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor samples
obtained from the Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}. See the
supplemental code\cite{sup} for examples of efficiently converting the
downloaded sample methylation data into an R matrix.

We removed 91,563 columns with all constant or missing values from the data,
leaving an $80\times 394014$ matrix. The example computes
pairs of methylation beta value column vectors with correlation values of 0.99
or more. We use the same {\tt tcor} projection dimension of 10 and performance
measurement methodology as used in the first example.

Algorithm \ref{prune} identified 913,601 pairs of vectors with correlation
values of 0.99 or more in about three hours on our test PC. A brute force
algorithm would need to evaluate over 155 billion vector pairs; the projection
of Algorithm \ref{prune} limited the search space to only about 4 million
candidate pairs, from which the 913,601 correlated pairs were found.

<<echo=FALSE, results=tex>>=
# Default values:
t1=10449.670
t2=NA
m1=1261.6
m2=NA

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  # simulated data, replicate rows 27x
}
require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99)", label = "TCGA3", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

@@

XXX

1. The main ideas in Algorithm \ref{prune}, although applied to correlation,
easily extend to any distance metric.

2. Anti-correlated pairs? Anyone?

@@

%% BROAD GDAC dashboard
\bibitem{gdac}{Broad Institute TCGA Genome Data Analysis Center (2015): Analysis-ready standardized TCGA data from Broad GDAC Firehose stddata\_\_2015\_06\_01 run. Broad Institute of MIT and Harvard. Dataset. \url{http://dx.doi.org/10.7908/C1251HBG}}

@@

%% IRLBA implementations and supplemental code
\bibitem{irlbar}
Lewis, Bryan,  \url{https://cran.r-project.org/web/packages/irlba}.

\bibitem{sup}
Lewis, Bryan,  \url{https://github.com/bwlewis/tcor}.
